---
title: "R Notebook"
output: html_notebook
---


```{r, error=TRUE, message=FALSE, warning=FALSE, purl=FALSE, results='hide'}
# Avoid duplicate label error of knitr::purl
options(knitr.duplicate.label = 'allow')
```

# Intro

First of all we are going to load required packages and the data. The data is part of the mlr3data package.


```{r message=FALSE, warning=FALSE}

if(!require("tidyverse")){install.packages("tidyverse")} # R packages for data science

```


# Downloading TCGA-KIRC clinical data from Xenabrowser


```{r}
# Survival data
# https://xenabrowser.net/datapages/

#url <- "https://gdc-hub.s3.us-east-1.amazonaws.com/download/TCGA-KIRC.survival.tsv"
destfile <- "../data/kirc_survival.tsv"
#download.file(url, destfile)
kirc_sur <- read_tsv(destfile)

kirc_sur <- kirc_sur %>% 
  column_to_rownames("sample") %>% 
  dplyr::rename(status = OS, obs.time = OS.time, patient_id = '_PATIENT')
  
#kirc_sur <- as.data.frame(kirc_sur)

# Transform days to months? / 30.43945

# Phenotype Clinical data
# https://xenabrowser.net/datapages/?dataset=TCGA-KIRC.GDC_phenotype.tsv&host=https%3A%2F%2Fgdc.xenahubs.net&removeHub=https%3A%2F%2Fxena.treehouse.gi.ucsc.edu%3A443
#url <- "https://gdc-hub.s3.us-east-1.amazonaws.com/download/TCGA-KIRC.GDC_phenotype.tsv.gz"
destfile <- "../data/kirc_clinical.tsv.gz"
#download.file(url, destfile)
kirc_cli <- read_tsv(gzfile(destfile))

kirc_cli <- kirc_cli %>%
  dplyr::select(c("submitter_id.samples","prior_malignancy.diagnoses","age_at_initial_pathologic_diagnosis", "gender.demographic", "race.demographic",
           "sample_type_id.samples", "pathologic_M", "pathologic_N", "pathologic_T")) %>% 
  dplyr::rename(sample = 'submitter_id.samples', 
         prior.dx = 'prior_malignancy.diagnoses', 
         age = 'age_at_initial_pathologic_diagnosis', 
         gender = 'gender.demographic',
         race = "race.demographic",
         sample.type = 'sample_type_id.samples',
         metastasis = 'pathologic_M',
         neoplasm = 'pathologic_N',
         ajcc.stage = 'pathologic_T') %>% 
  mutate(sample.type = str_replace_all(sample.type, "01", "TP") ) %>% 
  mutate(sample.type = str_replace_all(sample.type, "11", "NT") ) %>% 
  dplyr::filter(sample.type %in% c("TP", "NT")) %>%  
  dplyr::filter(sample %in% row.names(kirc_sur)) %>% 
  column_to_rownames("sample") %>% 
  mutate_if(is.character,as.factor)
  
kirc_cli <- cbind(kirc_cli, kirc_sur[rownames(kirc_cli),])

rm(kirc_sur)

```


# Downloading TCGA-KIRC counts from Xenabrowser

```{r}
# cohort: GDC TCGA Kidney Clear Cell Carcinoma (KIRC)
# https://xenabrowser.net/datapages/
       
#url <- "https://gdc-hub.s3.us-east-1.amazonaws.com/download/TCGA-KIRC.htseq_counts.tsv.gz"
destfile <- "../data/kirc_counts.tsv.gz"
#download.file(url, destfile)
kirc_rna <- read_tsv(gzfile(destfile))
kirc_rna <- as.data.frame(kirc_rna)
row.names(kirc_rna) <- kirc_rna$Ensembl_ID

#url <- "https://gdc-hub.s3.us-east-1.amazonaws.com/download/gencode.v22.annotation.gene.probeMap"
destfile <- "../data/kirc_probemap.tsv"
#download.file(url, destfile)
kirc_probe <- read_tsv(destfile)
kirc_probe <- as.data.frame(kirc_probe)
row.names(kirc_probe) <- kirc_probe$id
  
all(stringi::isUnique(kirc_probe$id))

kirc_rna  <- kirc_rna %>%
  dplyr::filter(Ensembl_ID %in% kirc_probe$id) %>%
  mutate(gene =  kirc_probe[Ensembl_ID, c("gene")])  %>%
  dplyr::distinct(gene, .keep_all = T) %>%
  dplyr::select(!c("Ensembl_ID")) %>%
  remove_rownames() %>%
  column_to_rownames("gene")

kirc_rna <- 2^(kirc_rna)-1
kirc_rna <- round(kirc_rna, digits = 0)

kirc_rna <- t(kirc_rna)



```

# Downloading TCGA-KIRC copy number (gene-level)

cohort: GDC TCGA Kidney Clear Cell Carcinoma (KIRC)  
https://xenabrowser.net/datapages/  

```{r}

#url <- "https://gdc-hub.s3.us-east-1.amazonaws.com/download/TCGA-KIRC.gistic.tsv.gz"
destfile <- "../data/kirc_gistic.tsv.gz"
#download.file(url, destfile)
kirc_gis <- read_tsv(gzfile(destfile))

kirc_gis <- as.data.frame(kirc_gis)

kirc_gis  <- kirc_gis %>%
  dplyr::filter(`Gene Symbol` %in% probeMap$id) %>%
  dplyr::select_if(!grepl("_1$", names(.))) %>%
  mutate(gene =  probeMap[`Gene Symbol`, c("gene")])  %>%
  dplyr::distinct(gene, .keep_all = T) %>%
  dplyr::select(!c("Gene Symbol")) %>%
  remove_rownames() %>%
  column_to_rownames("gene")

rm(list=setdiff(ls(), c("kirc_cli", "kirc_rna", "kirc_gis")))

```


# Filtering data for didactical purposes and simplify the binary metastasis Classification

```{r}

kirc_cli <- kirc_cli[kirc_cli$sample.type =="TP", ]

kirc_cli$sample.type <- NULL

kirc_cli$vital_status <- ifelse(kirc_cli$status ==0, "alive", "deceased")

# Remove patients without metastasis that died and with stage > T1
kirc_cli <- kirc_cli[kirc_cli$metastasis == "M1" | kirc_cli$vital_status =="alive", ]
kirc_cli <- kirc_cli[kirc_cli$metastasis == "M1" | (kirc_cli$ajcc.stage %in% c("T1","T1a", "T1b")), ]

rows <- (rownames(kirc_rna)  %in% rownames(kirc_cli))
kirc_rna <- kirc_rna[rows, ]

kirc_cli <- kirc_cli[rownames(kirc_rna), ]

kirc_cli <- kirc_cli %>% 
  distinct_at('patient_id', .keep_all = TRUE)   %>%
  dplyr::mutate(ajcc.stage = fct_collapse(ajcc.stage, # T1a: 118, T1b: 93, T3a: 40, T1: 16
                                          T1 = c('T1', 'T1a', 'T1b'),
                                          T2 = c('T2', 'T2b'),
                                          T3 = c('T3', 'T3a', 'T3b'),
                                          T4 = c('T4')),
                race = fct_recode(race,
                                  black='black or african american')) %>%
    droplevels()


rm(list=setdiff(ls(), c("kirc_cli", "kirc_rna")))

```



```{r}
#load("../data/tcga_kirc.RData")

kirc_cli$sample.type <- NULL

save(kirc_rna, kirc_cli, file="../data/tcga_kirc.RData", compress = T)

```